### 2025-02-16 src/paper/paper.tex

\tab\textbf{Background}\\
Schizophrenia (SZ) is a chronic mental disorder that presents challenges in accurate diagnosis due to its complex nature. Advances in machine learning (ML) and deep learning (DL) have shown promise in automating SZ detection using structural MRI, but variability in preprocessing methods remains a key challenge.
\textbf{Aim}: This study aims to evaluate the impact of various preprocessing and data augmentation techniques on ML-based classification of schizophrenia from MRI data.

@@

Among the individual preprocessing techniques, normalization resulted in the most substantial performance improvement, enhancing accuracy by 9.8\%. This can be attributed to the equalization of feature contributions, which allowed the model to make more accurate predictions. Interestingly, this outcome contradicts the conventional understanding that normalizing medical images may flatten clinically significant differences in intensity. The effectiveness of this technique in this study highlights the importance of context when applying preprocessing methods to medical imaging.

@@

However, combining preprocessing steps often led to diminished performance, especially when all techniques were applied together. The full preprocessing pipeline, including resampling, normalization, brain extraction, cropping, and smoothing, resulted in a performance drop, particularly in terms of accuracy and F1-score. This suggests that certain preprocessing steps, when combined, may interfere with each other, creating distortions that prevent the model from learning the most pertinent features of the data.

Regarding data augmentation, both translation and rotation significantly enhanced model performance, with translation improving recall and rotation achieving the highest accuracy and F1-score. These results align with the literature, where such transformations are often used to increase the diversity of the training data, helping the model generalize better to unseen data. Gaussian noise, however, severely degraded performance, suggesting that it may obscure important features in the MRI images, leading to poorer model predictions.

@@

Overall, this study highlights the importance of standardizing preprocessing and augmentation techniques in medical imaging, particularly for tasks like schizophrenia detection. While certain methods demonstrated clear benefits, the combination of multiple preprocessing steps should be approached with caution, as it may lead to diminishing returns or reduced model performance. Further research is needed to explore the optimal combination of preprocessing techniques and to investigate the impact of more advanced augmentation methods on model accuracy and generalizability.

@@

In this study, we investigated the impact of various preprocessing and data augmentation techniques on the classification performance of schizophrenia from structural MRI images. Our results demonstrate that individual preprocessing methods, particularly normalization, brain extraction, and cropping, can positively influence model performance, improving accuracy, recall, and F1-score. However, combining these techniques did not always lead to enhanced results, suggesting that over-processing may introduce distortions that hinder model generalization.

Data augmentation proved to be a valuable tool, with translation and rotation significantly improving model robustness and performance, particularly in recall and accuracy. Conversely, adding Gaussian noise resulted in a substantial performance decrease, indicating that it may obscure critical features in the data.

The findings emphasize the importance of careful selection and application of preprocessing and augmentation methods in medical imaging tasks. While some techniques show clear benefits, the combination of multiple preprocessing steps must be handled cautiously to avoid diminishing returns. These results underscore the need for a standardized preprocessing pipeline to enhance the reliability, reproducibility, and clinical applicability of machine learning models for schizophrenia detection.

Further work is needed to refine preprocessing pipelines and explore more advanced data augmentation methods to improve the performance of ML-based schizophrenia diagnosis. Ultimately, this study contributes to the ongoing efforts to make ML-driven medical diagnostics more robust and applicable in real-world clinical settings.

>>>>

\textbf{Background}\\
Schizophrenia (SZ) is a chronic mental disorder that presents challenges in accurate diagnosis due to its complex nature. Advances in machine learning (ML) and deep learning (DL) have shown promise in automating SZ detection using structural MRI, but variability in preprocessing methods remains a key challenge.
\textbf{Aim}: This study aims to evaluate the impact of various preprocessing and data augmentation techniques on ML-based classification of schizophrenia from MRI data.

@@

Among the individual preprocessing techniques, normalization resulted in the most substantial performance improvement, enhancing accuracy to 56\% (9.8\% increase compared to baseline). This can be attributed to the equalization of feature contributions, which allowed the model to make more accurate predictions. Interestingly, this outcome contradicts the conventional understanding that normalizing medical images may flatten clinically significant differences in intensity. The effectiveness of this technique in this study highlights the importance of context when applying preprocessing methods to medical imaging.

@@

However, combining preprocessing steps often led to diminished performance, especially when all techniques were applied together. While the best performance was achieved with resampling, normalization, and brain extraction, adding further steps, such as cropping and smoothing, resulted in a performance drop, particularly in terms of accuracy and F1-score. This suggests that certain preprocessing steps, when combined, may interfere with each other, creating distortions that prevent the model from learning the most pertinent features of the data.

Regarding data augmentation, both translation and rotation significantly enhanced model performance, with translation improving recall and rotation achieving the highest accuracy and F1-score. These results align with the literature, where such transformations are often used to increase the diversity of the training data, helping the model generalize better to unseen samples. Gaussian noise, however, severely degraded performance, suggesting that it may obscure important features in the MRI images, leading to poorer model predictions.

The final training model, which combined the optimal preprocessing steps (resampling, normalization, and brain extraction) with translation and rotation (1+2+3+6+7), achieved the highest overall performance, with an accuracy of 81\%, recall of 0.98, and an F1-score of 0.86. This result demonstrates the effectiveness of combining well-selected preprocessing and augmentation techniques in enhancing classification performance. Notably, the final model's recall score suggests that it is highly sensitive to positive cases, which is crucial for medical applications where detecting true positives is a priority.

@@

Overall, this study highlights the importance of standardizing preprocessing and augmentation techniques in medical imaging, particularly for tasks like schizophrenia detection. The results suggest that the combination of multiple preprocessing steps should be approached with caution, as it may lead to diminishing returns or reduced model performance while a well-optimized combination of preprocessing and augmentation can significantly improve model performance. Further research is needed to explore additional augmentation strategies and evaluate their impact on model generalizability in larger and more diverse datasets.

@@

This study examined the impact of preprocessing and data augmentation on schizophrenia classification from structural MRI images. The best preprocessing setup (resampling + normalization + brain extraction) improved performance (55\% accuracy, 0.69 F1-score), but excessive preprocessing did not yield further gains.

Data augmentation had a stronger effect, with rotation achieving the highest accuracy (79\%) and F1-score (0.80). The final model, trained with optimized preprocessing and augmentation, achieved the best overall performance (81\% accuracy, 0.98 recall), highlighting the value of spatial transformations.

The findings emphasize the importance of careful selection and application of preprocessing and augmentation methods in medical imaging tasks. The results underscore the need for a standardized preprocessing pipeline to enhance the reliability, reproducibility, and clinical applicability of machine learning models for schizophrenia detection.

Further work is needed to refine preprocessing pipelines and explore additional augmentation methods to improve the performance of ML-based schizophrenia diagnosis. Ultimately, this study contributes to the ongoing efforts to make ML-driven medical diagnostics more robust and applicable in real-world clinical settings.
